Iterative Monomorphisation

### 2024-03-04 iterative_monomorphisation.tex
some work with Tanguy
minor improvements on algos
some minor styling changes, still more to be done

\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage%[]
   {algorithm2e}
\usepackage{amsmath}

@@

\definecolor{grey}{rgb}{0.8,0.8,0.8}
\definecolor{algoColorKeyword}{named}{grey}
\makeatletter
% Block with a vertical line
\renewcommand{\algocf@Vsline}[1]{%
   \strut\par\nointerlineskip%
   \algocf@bblockcode%
   \algocf@push{\skiprule}%
   \hbox{{\color{algoColorKeyword}\vrule}%
      \vtop{\algocf@push{\skiptext}%
      \vtop{\algocf@addskiptotal\advance\hsize by -\skiplength #1}}%
   }%
   \algocf@pop{\skiprule}%
   \algocf@bblockcode%
}
\makeatother

@@

\begin{quote}
\begin{algorithm}[H]
   \caption{Substitution Generation}
%   \SetAlgoNoLine
%   \LinesNumbered
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKwData{Continue}{continue}\SetKwData{SubstRes}{substitution\_result}
   \SetKwData{SubstLimit}{substitution\_limit}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
   \Output{\,set of substitutions \(S\)}
   \BlankLine

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \Continue \(\leftarrow\) \True\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\) \And \Continue}{
            \SubstRes \(\leftarrow\) \Match{\(\upsilon_i\), \(\tau_i\)}\;
            \Case{\SubstRes = \Success{\(\sigma_i\)}}{
               \lIf{\Compatible{\(\sigma\), \(\sigma_i\)}}{
                  \(\sigma\leftarrow\sigma_i\circ\sigma\)}
               \lElse{\Continue\(\leftarrow\)\False}
            }
            \lOther{\Continue\(\leftarrow\)\False}

@@

         \If{\Continue}{
            \textcolor{BrickRed}{
               \If{\(|S| < \SubstLimit\)}{
                  \textcolor{black}{\(S\leftarrow S\cup \{\sigma\}\)\;}
               }
               \lElse{\Return \(S\)}
            }
         }
      }

   }
   \Return \(S\)\;
\end{algorithm}
\end{quote}

@@

\SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
\SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
\begin{quote}
\begin{algorithm}[H]
   \caption{Formula Monomorphisation Step}

@@

   \SetKwData{Subst}{substitutions}\SetKwData{TypeVars}{type\_variables}
   \SetKwData{MaxMono}{max\_mono\_args}
   \SetKwData{MaxPoly}{max\_poly\_args}
   \SetKwData{UsedSubst}{used\_subst}

@@

   \SetKwFunction{SubstGen}{substitution\_generation}\SetKwFunction{TyVars}{type\_variables}
   \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
   \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,old and new monomorphic type argument maps \(M, M'\), old and new non-monomorphic type argument maps \(P(\varphi), P'(\varphi)\)}
   \Output{\,monomorphic type argument map \(M''\), non-monomorphic type argument map \(P''(\varphi)\), set of substitutions \(S\)}
   \BlankLine

   \Subst \(\leftarrow\) \SubstGen{\(M'\), \(P'(\varphi)\)}\;
   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M'\), \(P(\varphi)\)}\;
   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P'(\varphi)\)}\;
   \BlankLine

@@

   \textcolor{BrickRed}{
   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M'| \times\)\MonoMult}, \MonoCap}\;
   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P'(\varphi)| \times\)\PolyMult}, \PolyCap}\;
   }

@@

   \UsedSubst \(\leftarrow\emptyset\)\;
   \ForEach{\(\sigma\in \Subst\)}{
      \ForEach{\(f\) occurring in \(\varphi\) }{
         \ForEach{\((\tau_1, \dots, \tau_n)\in P(\varphi)(f)\)}{
            \If{\TyVars{\((\tau_1, \dots, \tau_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
               \textcolor{BrickRed}{
               \If{\(|M''| < \MaxMono\) \label{algo:relative_bound}}{
                  \textcolor{black}{
                  \(M''(f)\leftarrow M''(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \ElseIf{\(|P''(\varphi)| \geq \MaxPoly\)}{
                  \Return \((M'', P''(\varphi), \UsedSubst)\)\;
               }}
            }
            \textcolor{BrickRed}{
            \ElseIf{\(|P''(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
               \textcolor{black}{
               \(P''(\varphi)(f)\leftarrow P''(\varphi)(f)\cup\{(\sigma(\tau_1),\dots,\sigma(\tau_n))\}\)\;
               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
               }
            }}

@@

   \Return \((M'', P''(\varphi), \UsedSubst)\)

\end{algorithm}
\end{quote}

@@

At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M''(f)\) (respectively \(P''(\varphi)(f)\)), would create finer grained bounds. This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated. Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.

\begin{quote}
\begin{algorithm}
   \caption{Monomorphising Substitutions}

   \SetKwFunction{MonoSubst}{monomorphising\_substs}
   \SetKwFunction{Domain}{domain}
   \SetKwData{Limit}{subst\_limit}
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \Input{\,set of substitutions \(U, S\), substitution \(\sigma\), set of type variables \(V\)}
   \Output{\,set of substitutions}
   \BlankLine

   \If{\(V = \emptyset\)}{
    %  \textcolor{BrickRed}{
    %  \If{\(|S|<\Limit\) }{
    %     \textcolor{black}{
    \Return \(S\cup\{\sigma\}\)\;%}
    %  }
    %  \lElse{\Return \(S\)}}
   }
   \Else{
      \(\exists\alpha\in V\)\;
      \ForEach{\(\sigma'\in U\) such that, \(\alpha\in\Domain{\(\sigma'\)}\) \And \(\sigma\) compatible with \(\sigma'\)}{
         \textcolor{BrickRed}{
         \If{\(|S|<\Limit\)}{
            \textcolor{black}{
            \(S\leftarrow \MonoSubst{U, S, \(\sigma'\circ\sigma\), \(V\setminus\Domain{\(\sigma'\)}\)}\)\;}
         }
         \lElse{\Return \(S\)}}
      }
      \Return \(S\)\;
   }

\end{algorithm}
\end{quote}

@@

\begin{quote}
\begin{algorithm}
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}

   \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{new\_formulae}
   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{used\_substitution}
   \SetKwData{Limit}{total\_new\_formulae}

   \caption{Iterative Monomorphisation}

   \((M, P) \leftarrow\) [empty maps]\;
   \((M', P') \leftarrow \Init{\Problem}\)\;

   \For{\(1\leq i\leq \textcolor{BrickRed}{\Loop}\) }{
      \(M''\leftarrow\emptyset\)\;
      \(P''\leftarrow\emptyset\)\;
      \ForEach{\(\varphi \in \Problem\)} {
         \((M^*, P^*(\varphi), S)\leftarrow \FStep(M, P(\varphi), M', P'(\varphi))\) \;
         \(\AllSubst\leftarrow\AllSubst\cup S\)\;
         \(M''\leftarrow M''\cup M^*\)\;
         \(P''(\varphi)\leftarrow P''(\varphi)\cup P^*(\varphi)\)\;
      }

      \((M, M')\leftarrow (M\cup M', M'')\)\;
      \((P, P')\leftarrow (P\cup P', P'')\)\;

   \NewFormulae \(\leftarrow\emptyset\)\;
   \For{\(\varphi\in\Phi\)}{
      \(V\leftarrow \TyVars{\(\varphi\)}\)\;
      \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
      \ForEach{\(\sigma\in S\)}{
         \textcolor{BrickRed}{
         \If{\(|\NewFormulae|<\Limit\)}{
            \textcolor{black}{
            \(\NewFormulae\leftarrow\NewFormulae\cup \sigma(\varphi)\)\;}
         }
         \lElse{\Return \NewFormulae}
      }}
   }
   \Return \NewFormulae\;

\end{algorithm}
\end{quote}

@@

During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \times \text{number of type arguments})) \)

>>>>

\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{array}
\usepackage{multirow}
\usepackage[noend]
   {algorithm2e}
\usepackage{amsmath}

@@

% algo2e package
\SetFuncSty{textsf}
\SetFuncArgSty{textit}
\SetArgSty{text}
\SetDataSty{textit}

\DontPrintSemicolon

\SetKwProg{Fn}{Function}{}{end}\SetKwFunction{FRecurs}{FnRecursive}%
\newcommand{\forcond}{$i=0$ \KwTo $n$}

\definecolor{ourblueviolet}{HTML}{0071BC}
\definecolor{grey}{rgb}{0.8,0.8,0.8}
\definecolor{algoColorKeyword}{named}{grey}
\makeatletter
% Block with a vertical line
\renewcommand{\algocf@Vsline}[1]{%
   \strut\par\nointerlineskip%
   \algocf@bblockcode%
   \algocf@push{\skiprule}%
   \hbox{{\color{algoColorKeyword}\vrule}%
      \vtop{\algocf@push{\skiptext}%
      \vtop{\algocf@addskiptotal\advance\hsize by -\skiplength #1}}%
   }%
   \algocf@pop{\skiprule}%
   \algocf@bblockcode%
}
\makeatother

@@

\SetKwFunction{IterMono}{iterative\_monomorphisation}
\begin{figure}
\begin{quote}
\begin{algorithm}[H]
\Fn(){\IterMono{\(\Phi\)}}{
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}

   \SetKwData{Loop}{loop\_nb}\SetKwData{NewFormulae}{\(\Psi\)}
   \SetKwData{Problem}{\(\Phi\)}\SetKwData{AllSubst}{\(U\)}
   \SetKwData{Limit}{total\_new\_formulae}

   \KwData{set of formulae \(\varphi\)}
   \KwResult{set of monomorphic formulae}

   \((M, P) \leftarrow (\emptyset, \emptyset)\)\;
   \((M^{\text{new}}, P^{\text{new}}) \leftarrow \Init{\Problem}\)\;

   \For{\(1\leq i\leq \textcolor{ourblueviolet}{\Loop}\) }{
      \((M^{\text{next}}, P^{\text{next}})\leftarrow(\emptyset,\emptyset)\)\;
      \ForEach{\(\varphi \in \Problem\)} {
         \((M^*, P^*(\varphi), S)\leftarrow \FStep(M, P(\varphi), M^{\text{new}}, P^{\text{new}}(\varphi))\)\;
         \(\AllSubst\leftarrow\AllSubst\cup S\)\;
         \(M^{\text{next}}\leftarrow M^{\text{next}}\cup M^*\)\;
         \(P^{\text{next}}(\varphi)\leftarrow P^{\text{next}}(\varphi)\cup P^*(\varphi)\)\;
      }

      \((M, M^{\text{new}})\leftarrow (M\cup M^{\text{new}}, M^{\text{next}})\)\;
      \((P, P^{\text{new}})\leftarrow (P\cup P^{\text{new}}, P^{\text{next}})\)\;

   \NewFormulae \(\leftarrow\emptyset\)\;
   \For{\(\varphi\in\Phi\)}{
      \(V\leftarrow \TyVars{\(\varphi\)}\)\;
      \(S\leftarrow\MonoSubst{\AllSubst, \(\emptyset\), \(\{\}\), \(V\)}\)\;
      \ForEach{\(\sigma\in S\)}{
         \textcolor{ourblueviolet}{
         \eIf{\(|\NewFormulae|<\Limit\)}{
            \textcolor{black}{
            \(\NewFormulae\leftarrow\NewFormulae\cup \{\sigma(\varphi)\}\)\;}
         }
         {\Return \NewFormulae}
      }}
   }
   \Return \NewFormulae\;

}
\end{algorithm}
\end{quote}
\caption{Pseudocode for iterative monomorphisation}
\end{figure}

\begin{figure}
\begin{quote}
\begin{algorithm}[H]
\SetKwFunction{SubstGen}{substitution\_generation}
\Fn(){\SubstGen{\(M, P(\varphi)\)}}{
%   \caption{Substitution Generation()}
%   \SetAlgoNoLine
%   \LinesNumbered
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKwData{SubstLimit}{substitution\_limit}
   \SetKwFunction{Success}{Success}\SetKwFunction{Compatible}{compatible}
   \SetKwFunction{Match}{match}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
   \SetKw{Break}{break}

   \KwData{monomorphic type argument map \(M\), non-monomorphic type argument map \(P(\varphi)\)}
   \KwResult{set of substitutions}
   \BlankLine

   \(S\leftarrow \emptyset\)\;
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\)}{
            \eIf{\(\Match{\(\upsilon_i\), \(\tau_i\)}\) has the form \Success{\(\sigma_i\)} \And \Compatible{\(\sigma\), \(\sigma_i\)}}{
               \(\sigma\leftarrow\sigma_i\circ\sigma\)
            }
            {\Break}

@@

         \If{\(i > n\)}{
            \textcolor{ourblueviolet}{
               \eIf{\(|S| < \SubstLimit\)}{
                  \textcolor{black}{\(S\leftarrow S\cup \{\sigma\}\)\;}
               }
               {\Return \(S\)}
            }
         }
      }

   }
   \Return \(S\)\;

}
\end{algorithm}
\end{quote}
\caption{Pseudocode for substitution generation}
\end{figure}

@@

\SetKwData{MonoCap}{mono\_cap}\SetKwData{MonoMult}{mono\_mult}\SetKwData{MonoFloor}{mono\_floor}
\SetKwData{PolyCap}{poly\_cap}\SetKwData{PolyMult}{poly\_mult}\SetKwData{PolyFloor}{poly\_floor}
\begin{figure}
\begin{quote}
\begin{algorithm}[H]

@@

   \SetKwData{Subst}{substitutions}\SetKwData{TypeVars}{type\_variables}
   \SetKwData{MaxMono}{max\_mono\_args}
   \SetKwData{MaxPoly}{max\_poly\_args}
   \SetKwData{UsedSubst}{\(U\)}

@@

   \SetKwFunction{SubstGen}{substitution\_generation}\SetKwFunction{TyVars}{type\_variables}
   \SetKwFunction{Max}{max}\SetKwFunction{Min}{min}
   \SetKwFunction{Domain}{subst\_domain}\SetKwFunction{Len}{len}

   \KwData{old and new monomorphic type argument maps \(M, M^{\text{new}}\), old and new non-monomorphic type argument maps \(P(\varphi), P^{\text{new}}(\varphi)\)}
   \KwResult{monomorphic type argument map \(M^{\text{next}}\), non-monomorphic type argument map \(P^{\text{next}}(\varphi)\), set of substitutions \(S\)}
   \BlankLine

   \Subst \(\leftarrow\) \SubstGen{\(M^{\text{new}}\), \(P^{\text{new}}(\varphi)\)}\;
   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M^{\text{new}}\), \(P(\varphi)\)}\;
   \Subst \(\leftarrow\) \Subst \(\cup\) \SubstGen{\(M\), \(P^{\text{new}}(\varphi)\)}\;
   \BlankLine

@@

   \textcolor{ourblueviolet}{
   \MaxMono \(\leftarrow\) \Max{\Min{\MonoFloor, \(|M\cup M^{\text{new}}| \cdot \)\MonoMult}, \MonoCap}\;
   \MaxPoly \(\leftarrow\) \Max{\Min{\PolyFloor, \(|P(\varphi)\cup P^{\text{new}}(\varphi)| \cdot \)\PolyMult}, \PolyCap}\;
   }

@@

   \UsedSubst \(\leftarrow\emptyset\)\;
   \ForEach{\(\sigma\in \Subst\)}{
      \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in P(\varphi)\)}{
            \eIf{\TyVars{\((\upsilon_1, \dots, \upsilon_n)\)} \(\subseteq\) \Domain{\(\sigma\)}}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M^{\text{next}}| < \MaxMono\) \label{algo:relative_bound}}{
                  \textcolor{black}{
                  \(M^{\text{next}}(f)\leftarrow M^{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|P^{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  blablabliblou\;
                  \Return \((M^{\text{next}}, P^{\text{next}}(\varphi), \UsedSubst)\)\;
               }}
            }
            {
               \textcolor{ourblueviolet}{
               \If{\(|P^{\text{next}}(\varphi)| < \MaxPoly\)\label{algo:ref_2}} {
               \textcolor{black}{
               \(P^{\text{next}}(\varphi)(f)\leftarrow P^{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
               \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;
            }}
            }

@@

   \Return \((M^{\text{next}}, P^{\text{next}}(\varphi), \UsedSubst)\)

\end{algorithm}
\end{quote}
\caption{Formula Monomorphisation Step}
\end{figure}

@@

At line \ref{algo:relative_bound} (and similarly at line \ref{algo:ref_2}), adding conditions dependent on the size of \(M^{\text{next}}(f)\) (respectively \(P^{\text{next}}(\varphi)(f)\)), would create finer grained bounds. This may be useful if some function symbols have too many type arguments and prevent new type arguments of other function symbols from being generated. Adding a cap on the number of new type arguments per function symbol may help in smoothing out the distribution of new type arguments over function symbols.

\begin{figure}
\begin{quote}
   \begin{algorithm}[H]

   \SetKwFunction{MonoSubst}{monomorphising\_substs}
   \SetKwFunction{Domain}{domain}
   \SetKwData{Limit}{subst\_limit}
   \SetKw{And}{and}\SetKw{True}{true}\SetKw{False}{false}\SetKw{Stop}{stop}
   \SetKw{Let}{let}
   \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

   \KwData{set of substitutions \(U\) and \(S\), single substitution \(\sigma\), set of type variables \(V\)}
   \KwResult{set of substitutions}
   \BlankLine

   \eIf{\(V = \emptyset\)}{
    \Return \(S\cup\{\sigma\}\)\;%}
   }
   {
      \Let \(\alpha \) such that \(\alpha\in V\)\;
      \ForEach{\(\sigma'\in U\) such that, \(\alpha\in\Domain{\(\sigma'\)}\) \And \(\sigma\) compatible with \(\sigma'\)}{
         \textcolor{ourblueviolet}{
         \eIf{\(|S|<\Limit\)}{
            \textcolor{black}{
            \(S\leftarrow \MonoSubst{U, S, \(\sigma'\circ\sigma\), \(V\setminus\Domain{\(\sigma'\)}\)}\)\;}
         }
         {\Return \(S\)}}
      }
      \Return \(S\)\;
   }

\end{algorithm}
\end{quote}
\caption{Monomorphising Substitutions}
\end{figure}

@@

@@

During an iteration, for each symbol (for each clause in the polymorphic case), the number of newly generated type arguments is determined by this formula:
\( \min(\text{cap}, \max(\text{floor}, \text{multiplier} \cdot \text{number of type arguments})) \)
